---
title: "NYPD_week_3"
output: html_document
date: "2023-03-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Project Description**
The project will import NYC shooting data from (https://catalog.data.gov/dataset?q=NYPD+Shooting+Incident+Data+%28Historic%29) and analyze it. The data breaks down shooting victims and perpetrators by borough, race, sex, age, location, and indicates whether or not the shooting was a murder. 

1) What are the murdered victims by race? Sex? Age? 
2) In what borough are the most murdered victims? 

In this analysis, I will group the data by murdered victims by borough, race, sex, and age.


```{r}
library(tidyverse)
library(lubridate)
```

## **Importing Data** (Step 1)
```{r}
data_in <- read_csv("NYPD_Shooting_Incident_Data__Historic_.csv")
colnames(data_in)
summary(data_in)
```

## **Tidy up, Transform, and Handle the Missing Data** (Step 2)
We observe that there are some missing values for variables LOCATION_DESC, PERP_AGE_GROUP, PERP_SEX, and PERP_RACE. Since it does not make sense to assume values for those missing variables, we will remove them entirely. Another justification for not considering the specified variables is that, for this project, we are interested solely in the statistics about victims and where shooting incidents were more prevalent in the city. For that reason, we are also not interested in variables INCIDENT_KEY, OCCUR_TIME, JURISDICTION_CODE, PRECINCT, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, and Lon_Lat. This transforms the data frame. We will also convert OCCUR_DATE to a date object from a character string and finally show a summary of the data. 

```{r}
data_in <- data_in %>% select(-c(INCIDENT_KEY, OCCUR_TIME, JURISDICTION_CODE, PRECINCT, LOCATION_DESC, PERP_AGE_GROUP, PERP_SEX, PERP_RACE, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat)) %>% mutate(OCCUR_DATE=mdy(OCCUR_DATE))
summary(data_in)
```

## **Analyze the Data** (Step 3a)
We will perform two different analyses of the data. 1) We want to know the murder counts by borough. We will group the data by BORO and summarize the murder counts. 2) We are also interested in the murdered vicitms by race in the city as a whole. 

```{r}
murders_by_borough <- data_in %>% group_by(BORO) %>% summarise(TOTAL_MURDERS = sum(STATISTICAL_MURDER_FLAG == TRUE))
victims_by_race <- data_in %>% group_by(VIC_RACE) %>% summarise(TOTAL_MURDERS = sum(STATISTICAL_MURDER_FLAG == TRUE))
victims_by_age <- data_in %>% group_by(VIC_AGE_GROUP) %>% summarise(TOTAL_MURDERS = sum(STATISTICAL_MURDER_FLAG == TRUE))
victims_by_sex <- data_in %>% group_by(VIC_SEX) %>% summarise(TOTAL_MURDERS = sum(STATISTICAL_MURDER_FLAG == TRUE))
```

## **Visualize the Data** (Step 3b)
We want to visualize both the total murders by borough as well as the murder counts by race using bar graphs. 
```{r, echo=FALSE}
murders_by_borough %>% ggplot(aes(x=BORO, y=TOTAL_MURDERS)) + geom_bar(stat="identity", width=0.1, color="blue") + labs(title="Murder Counts by Borough", y=NULL) + theme(plot.title = element_text(hjust=0.5))
victims_by_race %>% ggplot(aes(x=VIC_RACE, y=TOTAL_MURDERS)) + geom_bar(stat="identity", width=0.1, color="red") + theme(axis.text.x=element_text(angle=90)) + labs(title="Murder Counts by Victim's Race", y=NULL) + theme(plot.title = element_text(hjust=0.5))
victims_by_age %>% ggplot(aes(x=VIC_AGE_GROUP, y=TOTAL_MURDERS)) + geom_bar(stat="identity", width=0.1, color="green") + labs(title="Murder Counts by Victim's Age Group", y=NULL) + theme(plot.title = element_text(hjust=0.5))
victims_by_sex %>% ggplot(aes(x=VIC_SEX, y=TOTAL_MURDERS)) + geom_bar(stat="identity", width=0.1, color="yellow") + labs(title="Murder Counts by Victim's Sex", y=NULL) + theme(plot.title = element_text(hjust=0.5))
```

The data has determined that the most murders by borough occurred in Brooklyn, with the most murdered victims by race being black, by sex being male, and by age group between 25-44. Though no assumptions about the data should be made at this point, it can lead to the posing of questions such as: why are there so many murdered victims in Brooklyn? Is it solely due to population, or are there other reasons? Another potential question to be investigated is why are most of the murdered victims black, and how that can change based on the borough. 


## **Bias Identification and Elimination** (Step 4)
In conclusion, it was determined that the most murders by borough occurred in Brooklyn, and the most murdered victims were Black by race, male by sex, and 25-44 by age. 
By removing the variables PERP_AGE_GROUP, PERP_SEX, and PERP_RACE instead of filling in arbitrary values for the missing values in those columns, we have eliminated sources of bias about what groups of people have been carrying out the shootings. I assumed that the most populated borough would have the most murders, which turned out to be the case with Brooklyn. However, it cannot be assumed that Brooklyn has the most cases solely because of population, and more investigation is needed. 